---
title: "NHSStepUpReport"
output: html_document
date: "2025-03-05"
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(scales)
library(gghighlight)
library(ggthemes)
library(rgeoboundaries)
library(sf)
library(ggmap)
df <- read_rds(file = "STEP_UP_REGIONAL_ANTIDEPRESSANTS.Rds")
```

# Part One

**1** *Create two vertical bar charts for comparison. First, create a
bar chart showing the total annual antidepressant prescribing (items).
Second, create a bar chart that shows the total antidepressant
prescribing cost. Describe the trend of each chart and compare the
charts against each other.*

```{r Part 1, two bar charts}
b1 <- df %>% 
  group_by(YEAR) %>% 
  summarise(totalCOUNT = sum(ITEMS)) %>% 
  ungroup() %>% 
  ggplot(aes(x = YEAR, y = totalCOUNT, fill = YEAR)) + 
  geom_bar(stat = "identity") + 
  labs(x = 'Year',
       y = 'Total Number of Prescriptions') + 
  scale_y_continuous(labels = label_comma()) + 
  theme(legend.position = 'none')
  
b2 <- df %>% 
  group_by(YEAR) %>% 
  summarise(totalCOST = sum(COST)) %>% 
  ungroup() %>% 
  ggplot(aes(x = YEAR, y = totalCOST, fill = YEAR)) + 
  geom_bar(stat = "identity") + 
  labs(x = 'Year', y = 'Total Cost of Prescriptions') + 
  scale_y_continuous(labels = label_comma()) + 
  theme(legend.position = 'none')
  
b1
b2
```

**2** *Create two tables for comparison. First, create a table showing
the total annual antidepressant prescribing per region (items,
pivot_wider). Second, create a table showing the annual antidepressant
prescribing cost per region (pivot_wider). Describe some regional
changes and contrasts between 2021 and 2024. (NOTE: within mutate a
numerical column name, e.g. 2021 must be placed with ticks, so 2021, for
example mutate(NEW_VAL = 2021 + 2022))*

```{r}
t1 <- df %>% 
  group_by(REGION, YEAR) %>% 
  summarise(totalCount = sum(ITEMS)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = REGION, values_from = totalCount)

t1

t2 <- df %>% 
  group_by(REGION, YEAR) %>% 
  summarise(totalCost = sum(COST)) %>% 
  ungroup() %>% 
  pivot_wider(names_from = REGION, values_from = totalCost)

t2
```

Summary:

From 2021 to 2024, overall costs have significantly decreased
nationally - likely with 2021 as an outlier from COVID. Total
prescriptions have gone up steadily since 2021 across all regions, with
some increasing more tan others i.e NE and Yorkshire. Total costs from
2023 \> 2024 have decreased in most regions, but increased in London.

**3** *By now you may have noticed there are contrasts between
antidepressant prescribing volumes and costs. Next, we will create two
horizontal bar charts for comparison. First, create an ordered bar chart
showing the 10 most prescribed antidepressants in the most recent year
(items). Second, create an ordered bar chart that shows antidepressants
with the greatest total prescribing cost in the most recent year.
Highlight and describe any differences between the charts.*

```{r}
b3_count <- df %>% 
  group_by(DRUG) %>% 
  summarise(totalCount = sum(ITEMS)) %>% 
  ungroup() %>% 
  top_n(10, totalCount) %>% 
  mutate(DRUG = fct_reorder(DRUG, totalCount)) %>% 
  ggplot(aes(x = DRUG, y = totalCount, fill = DRUG)) +  
  geom_col() + 
  scale_y_continuous(labels = label_comma()) +  
  coord_flip() +  
  theme_minimal() +
  labs(title = "Top 10 Drugs by Total Count",
       x = "Drug",
       y = "Total Count") + 
  theme(legend.position = 'none')

b3_cost <- df %>% 
  group_by(DRUG) %>% 
  summarise(totalCost = sum(COST)) %>% 
  ungroup() %>% 
  top_n(10, totalCost) %>% 
  mutate(DRUG = fct_reorder(DRUG, totalCost)) %>%  
  ggplot(aes(x = DRUG, y = totalCost, fill = DRUG)) +  
  geom_col() + 
  scale_y_continuous(labels = label_comma()) +  
  coord_flip() +  
  theme_minimal() +
  labs(title = "Top 10 Drugs by Total Cost",
       x = "Drug",
       y = "Total Cost") + 
  theme(legend.position = 'none') + 
  theme_economist(base_size = 7) + 
  theme(legend.position = 'none') 
  
b3_count
b3_cost
```

```{r}
df_combined <- df %>% 
  group_by(DRUG) %>% 
  summarise(totalCount = sum(ITEMS), 
            totalCost = sum(COST)) %>% 
  ungroup() %>% 
  top_n(10, totalCount) %>% 
  pivot_longer(cols = c(totalCount, totalCost), names_to = "CostOrCount", values_to = "values") %>% 
  mutate(DRUG = fct_reorder(DRUG, values))

b_facet <- df_combined %>% 
  ggplot(aes(x = DRUG, y = values, fill = DRUG)) +
  geom_col() + 
  coord_flip() + 
  facet_wrap(~CostOrCount, scales = "fixed") + 
  scale_y_continuous(labels = label_comma()) + 
  labs(x = 'Drug', 
        y = "Cost, Volume of Drug", 
        title = 'Total Cost vs Total Count of Drug') + 
  theme_economist(base_size = 7) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position = 'none')

b_facet
```

```{r}
cbar <- df_combined %>% 
  ggplot(aes(x = DRUG, y = values, fill = CostOrCount)) + 
  geom_col() + 
  scale_y_continuous(labels = label_comma()) + 
  theme_economist(base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = 'Drug', 
        y = "Cost, Volume of Drug", 
        title = 'Total Cost vs Total Count of Drug', 
       fill = "Metric")

cbar
```

# Part Two

*Now you have a good understanding around national and regional
antidepressant prescribing volume and cost trends. You will be given
various topics or points to explore, and you will decide what approach
or charts best does this. To supplement the initial analysis, you will
now:*

**Look at the longitudinal (monthly) trend for all antidepressant items
and for the cost of antidepressant prescribing**

```{r}
p2df <- df %>% 
  group_by(YM) %>% 
  summarise(totalCost = sum(COST),
            totalCount = sum(ITEMS)) %>% 
  ungroup() %>%
  pivot_longer(cols = c(totalCost, totalCount), names_to = "Metric", values_to = "Value")

p2df %>% 
  ggplot() + 
  geom_line(aes(x = YM, y = Value, colour = Metric, group = Metric)) + 
  theme_economist(
    base_size = 7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs (x = 'Time')
  
```

**Then, explore the data and find the antidepressant drugs that are
driving these monthly item and cost trends.**

```{r}
m1 <- df %>% 
  group_by(DRUG, YM) %>% 
  summarise(DrugCost = sum(COST)) %>% 
  ungroup() %>% 
  group_by(YM) %>% 
  top_n(10, DrugCost) %>% 
  ungroup() %>% 
  mutate(YM = as.numeric(YM),
         StartDate = StartDate <- as.Date("2021-01-01"),
         YearCount = YearCount <- StartDate + months(YM))

m1 %>% 
  ggplot() + 
  geom_line(aes(YearCount, DrugCost, color = DRUG, group = DRUG)) +
  gghighlight(max(DrugCost) > 2750000, use_direct_label = TRUE) + 
  scale_y_continuous(labels = label_comma()) +
  labs(y = "Cost of Drug",
       x = 'Years', 
       title = 'Cost of Drugs since 2021', 
       subtitle = 'Top 10 highest costing drugs',
       caption = 'Highlighted highest costed drugs') + 
  theme_economist() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 5))) + 
  theme(plot.subtitle = element_text(hjust = 0.5, size = 10))
```

```{r}
m2 <- df %>% 
  group_by(DRUG, YM) %>% 
  summarise(Costs = sum(COST),
            Totals = sum(ITEMS)) %>% 
  ungroup() %>% 
  group_by(YM) %>% 
  pivot_longer(cols = c(Costs, Totals), names_to = "Metric", values_to = "Value") %>% 
  ungroup() %>% 
  mutate(YM = as.numeric(YM),
         StartDate = StartDate <- as.Date("2021-01-01"),
         YearCount = YearCount <- StartDate + months(YM)) %>% 
  group_by(DRUG, YM) %>% 
  top_n(3, Value)

m2 %>% ggplot() +
  geom_line(aes(x = YearCount, y = Value, color = DRUG, group = interaction(DRUG, Metric), linetype = Metric)) +
  scale_y_continuous(labels = label_comma()) + 
  theme_economist(base_size = 7) + 
  theme(legend.position = 'none') +
  labs(x = 'Years',
       title = 'Drug cost and prescriptions over time',
       subtitle = 'since 2021') + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 5))) + 
  theme(plot.subtitle = element_text(hjust = 0.5, size = 12))
```

```{r}
m4 <- df %>% 
  group_by(YM) %>% 
  summarise(AvgCost = mean(COST),
            AvgTotal = mean(ITEMS)) %>% 
  ungroup() %>% 
  group_by(YM) %>% 
  pivot_longer(cols = c(AvgCost, AvgTotal), names_to = "Metric", values_to = "Value") %>% 
  ungroup() %>% 
  mutate(YM = as.numeric(YM),
         StartDate = StartDate <- as.Date("2021-01-01"),
         YearCount = YearCount <- StartDate + months(YM))

m4 %>% ggplot() + 
  geom_line(aes(x = YearCount, y = Value, linetype = Metric, color = Metric)) + 
  theme_economist() + 
  labs(title = 'Average Cost and Prescriptions of Drugs over time',
       x = 'Years',) + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 10)))
```

**Then give a high-level summary to your work and findings.**

The large increase that was found in 2021 to overall drug cost seems to
be primarily due to one drug: sertraline hydrochloride. Other than this
drug, the cost of drugs in 2021 remains on par with other drugs and
other years going foward. Overall then, we see a fairly stable cost of
drug over time, with some spikes being found in citalopram hydrobromide
and setraline hydrochloride in mid-20222 and mid-2024 respectively. The
reason for these would need to be investigated using real-world context,
rather than the data. The mean items vs costs over time clearly shows
the distinction that costs have incresaed despite no increase in items
prescribed. There is another mediating factor at play.

**Part Two Extension (Antidepressant Case studies)**

Focus one one or more antidepressants that play a significant role in
national volume and cost trends. Sertraline hydrochloride, citalopram
hyrdrobromide, Venlafaxine

Try to offer more granular insight into their effect on national
prescribing trends Percentage of total antidepressant volume or cost
from a drug

```{r}
prop1 <- df %>% 
  mutate(YM = as.numeric(YM),
         StartDate = StartDate <- as.Date("2021-01-01"),
         YearCount = YearCount <- StartDate + months(YM)) %>% 
  group_by(YearCount) %>% 
  mutate(SertCount = sum(ITEMS[DRUG == "Sertraline hydrochloride"]),
         VenCount = sum(ITEMS[DRUG == "Venlafaxine"]),
         CetaCount = sum(ITEMS[DRUG == "citalopram hyrdrobromide"]),
         TotalItems = sum(ITEMS),
         SertProp = SertCount / TotalItems,
         VenProp = VenCount / TotalItems,
         CetaProp = CetaCount / TotalItems) %>% 
  pivot_longer(cols = c(SertProp, VenProp, CetaProp), names_to = "Drug", values_to = "Proportion")

prop1 %>% ggplot() + 
  geom_line(aes(x = YearCount, y = Proportion, color = Drug)) + 
  scale_y_continuous(labels = label_percent()) +
  labs(x = 'Years', 
       y = 'Proporiton of Total Items Prescribed', 
       title = 'Proprotion of Total Annual Prescriptions from selected drugs', 
       caption = 'Selected Drugs') + 
  scale_colour_discrete(labels = c("SertProp" = "Sertraline", "CetaProp" = "Citalopram", "VenProp" = "Venlafaxine")) +
  theme_economist() + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 10)))
```

```{r}
prop2 <- df %>% 
  mutate(YM = as.numeric(YM),
         StartDate = StartDate <- as.Date("2021-01-01"),
         YearCount = YearCount <- StartDate + months(YM)) %>% 
  group_by(YearCount) %>% 
  mutate(SertCount = sum(ITEMS[DRUG == "Sertraline hydrochloride"]),
         VenCount = sum(ITEMS[DRUG == "Venlafaxine"]),
         CetaCount = sum(ITEMS[DRUG == "citalopram hyrdrobromide"]),
         SertCost = sum(COST[DRUG == "Sertraline hydrochloride"]),
         VenCost = sum(COST[DRUG == "Venlafaxine"]),
         CetaCost = sum(COST[DRUG == "citalopram hyrdrobromide"]),
         TotalItems = sum(ITEMS),
         TotalCost = sum(COST),
         SertProp = SertCount / TotalItems,
         VenProp = VenCount / TotalItems,
         CetaProp = CetaCount / TotalItems,
         SertCostProp = SertCost / TotalCost,
         VenCostProp = VenCost / TotalCost,
         CetaCostProp = CetaCost / TotalCost) %>% 
  pivot_longer(cols = c(SertProp, VenProp, CetaProp, SertCostProp, VenCostProp, CetaCostProp), names_to = "Drug", values_to = "Proportion") %>% 
  mutate(Type = if_else(str_detect(Drug, "CostProp"), "CostProp", "ItemProp"))
  
prop2 %>% ggplot(aes(x = YearCount, y = Proportion, colour = Drug, linetype = Type)) + 
  geom_line() + 
  gghighlight(str_detect(Drug, "Prop"))
  

```

**Mean cost per item in relation a drug; Distribution of mean cost per
item values for a drug**

```{r}
df %>% 
  group_by(DRUG) %>% 
  summarise(ItemTotal = sum(ITEMS),
            ItemCost = sum(COST),
            CperI = ItemCost/ItemTotal) %>% 
  top_n(7, CperI) %>% 
  mutate(DRUG = fct_reorder(DRUG, CperI, .desc = TRUE)) %>% 
  ggplot(aes(x = DRUG, y = CperI, fill = DRUG)) + 
  geom_col() + 
  theme_economist() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position = 'none') + 
  scale_y_continuous(labels = function(x) paste0("£", x)) + 
  labs(y = 'Cost per Item',
       x = 'Drug',
       title = 'Cost of Individual Drugs, per Item',
       caption = 'Top 7 highest cost per item drugs') + 
  theme(plot.title = element_text(hjust = 0.5, size = 14, margin = margin(b = 10, t = 5))) + 
  theme(axis.title.y = element_text(hjust = 0.5, size = 12, margin = margin(r = 5)))

```


```{r}
uk <- gb_adm0("GBR")

latitudes <- c(52.252929, 51.5072, 52.621223, 54.526905, 54.099752, 51.129932, 51.039854)
longitudes <- c(0.710673, -0.1276, -1.389856,-1.564316, -2.607994, 0.474771, -2.746780)

drug_map_data <- df %>% 
  group_by(REGION) %>% 
  summarise(itemTotals = sum(ITEMS)) %>% 
  ungroup()

drug_map_data <- cbind(drug_map_data, y_lat = latitudes, x_long = longitudes)
  
p <- ggplot() + 
  geom_sf(data = uk) +
  geom_point(data = drug_map_data, aes(x = x_long, y = y_lat, size = itemTotals, color = REGION, alpha = .3)) + 
  scale_size_continuous(range = c(10, 25)) +
  scale_color_discrete(expression(paste(bold(name = 'Region')))) +
  ylim(50, 55) +
  guides(size = 'none', alpha = 'none') + 
  theme_void() + 
  labs(title = expression(paste(bold('Map of Drug Prescriptions by Region'))), 
       caption = expression(paste(italic('*Not adjusted for region size and population'))),
       subtitle = 'The Midlands and North East and Yorkshire have the largest proporition of
       drug prescriptons in the last years (2021-2014).') + 
  theme(plot.title = element_text(hjust = 0, size = 18, margin = margin(t = 20))) + 
  theme(plot.caption = element_text(hjust = 1, size = 10, margin = margin(b = 10))) + 
  theme(plot.subtitle = element_text(hjust = 0, size = 10, margin = margin(b = 15)))
  
p
```
Note: when scaled for HTML, the above map of regional prescriptions does not scale the best - if outputed independantly it can be correctly scaled.

**Then give a high-level summary to your work and findings.**
The previous analyses show that despite Sertraline hydrochloride being the most prescribed drug, account for 24% of all drug prescriptions nationally, it is not the most cost intensive. 
Per item, it doesn't crack the top 7. Tranylcypromine sulfate is the highest cost per item drug, followed by nefazodone hyrdrochloride, however the difference between the two is around £500. 
By region, the North East and Yorkshire and Midlands account for the most amount of drug items prescribed, however that does not account for region size and relative population, so those figures and visualisations must be interpreted knowing such. As previously mentioned, the and elaborted on by examining individual drugs and their proportions of total item count and cost, prescriptions and costs are not necessarily related both across all drugs and these drugs individually. Whilst individual drugs do fluctuate in terms of cost, e.g sertraline, their volume stays stable throughout the past 4 years.

